---
title: "Activity Monitoring"
author: "Nathi"
output: html_document
---
The following data analysis will be carried out on data obtained from an individual's
personal activity monitoring device. The data can be obtained [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).  

## 1. Loading and Reading the Data

### Downloading the Data
The data can be downloaded to the desired directory using the following R commands:
 ```{r Downloading, cache=TRUE}
 if(!file.exists("./Data")){dir.create("./Data")}
 fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
 download.file(fileURL, destfile = "./Data/activity.zip")
 ```
 
### Reading the Data
 The function `read.csv()` is then used to read the data file and store it in the variable *activityData*
```{r Reading}
activityData <- read.csv("./Data/activity.csv", stringsAsFactors = FALSE)
```

Lets take a look at the data using `str()` function
```{r}
str(activityData)
```

The class of the **date** variable is set as character. It would be more convenient to convert it to 
a date class using `as.Date()` function as follows:
```{r}
activityData$date <- as.Date(activityData$date, "%Y-%m-%d")
```
## 2. Calculating the Mean Total Number of Steps Taken Per Day

### Calculating the Total Steps Per Day
In order to calculate the total steps per day, the variable *steps* has to be grouped by the variable *date* using the `group_by()`. To find the sum, `summarise()` function is used as shown. These function are available in the dplyr package.
```{r Sum, warning=FALSE, message=FALSE}
library(dplyr)
totalSteps <- activityData %>% group_by(date) %>% summarise(Total.Steps = sum(steps, na.rm = TRUE))
head(totalSteps)
```