---
title: "Activity Monitoring"
author: "Nathi"
output: html_document
---
The following data analysis will be carried out on data obtained from an individual's
personal activity monitoring device. The data can be obtained [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).  

## 1. Loading and Reading the Data

### Downloading the Data
The data can be downloaded to the desired directory using the following R commands:
 ```{r Downloading, cache=TRUE}
 if(!file.exists("./Data")){dir.create("./Data")}
 fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
 download.file(fileURL, destfile = "./Data/activity.zip")
 ```

### Reading the Data
 The function `read.csv()` is then used to read the data file and store it in the variable *activityData*
```{r Reading, cache=TRUE}
activityData <- read.csv("./Data/activity.csv", stringsAsFactors = FALSE)
```

Lets take a look at the data using `str()` function
```{r}
str(activityData)
```

The class of the **date** variable is set as character. It would be more convenient to convert it to 
a date class using `as.Date()` function as follows:
```{r}
activityData$date <- as.Date(activityData$date, "%Y%m%d")
```